My wonderful paper

### 2024-09-13 _manuscript/_tex/index.tex
more definition and explanation on expectations and the current gap

\begin{itemize}
\tightlist
\item
  describe/ define what analysis plan is, expectation (outcome and plan)

  \begin{itemize}
  \tightlist
  \item
    expectations can be used to formulate unit tests, they can be used
    to guide to diagnose in the analysis process: data issue,
    unreasonable expectation -- expectation is ``wrong''
  \item
    some expectations can be about data: e.g.~check temporal gaps in
    time series data. Some others are statistical checks: sample size,
    outliers, model specification, or checks based on domain knowledge:
    include temperature in the mortality \textasciitilde{} PM10 model.
  \end{itemize}
\end{itemize}

@@

An analysis plan is a set of analysis steps plus their expected
outcomes. An analysis plan consists of two parts: outcome expectation
and plan expectation. Outcome expectation is the expected result of the
analysis, for example, a significant p-value or a coefficient estimate
within a certain range. Plan expectation is the expected results from
the intermediate steps of the analysis, for example, the expected
distribution of the data, or the expected correlation structure between
variables.

While there is typically one major result expectation in an analysis,
multiple plan expectations can be formulated as unit tests to guide the
analysis process. These plan expectations serve as road marks to test,
at various stages of the analysis, whether the results are within
expectation. When the outcome is out of expectation, these plan
expectations can serve as unit tests to diagnose the analysis process.

@@

  \begin{itemize}
  \tightlist
  \item
    with simulation, the unit tests can be used to construct the
    ``result universe''
  \item
    With formulated expectation, one can generate multiple version of
    the dataset for teaching statistics
  \end{itemize}
\end{itemize}

>>>>

@@

TODO: more about analysis plan itself here

TODO: what if the expectation is ``wrong''

An analysis plan is a set of analysis steps combined with expectations.
Expectations represent our belief about certain aspects of the analysis,
independent of the analysis itself. It can be divided into two types:
\emph{outcome expectation} and \emph{plan expectation}. Outcome
expectation refers to what we anticipate from the main result of the
analysis based on prior knowledge. They shape how we interpret the
results and assess whether they are consistent with existing knowledge
or indicate the need for updates \citep{grolemund_cognitive_2014}. For
example, in public health, prior research shows the average increase in
mortality rate per unit increase in PM10 is about 0.50\%
\citep{liu2019ambient}. This serves as an expectation for similar future
studies. Plan expectations concern the intermediate steps within the
analysis rather than the final outcome. They serve as checkpoints to
detect deflection in the analysis process For example, we may expect
temporal data to be ordered with no gaps and duplicates, or expect that
temperature will be a significant covariate in the linear regression
model of PM10 on mortality.

(might be useful) Analysis plans can be constructed at various
granularities, at the highest level, one may only has a plan of the
specific method used for analysing data and the expected outcome. This
provides little guidance when a deviation from expectation occurs. At
the lowest level, one may have a plan for each data entry and every data
handling steps. This provides too much detail and may not be practical
in practice.

Experienced analysts often have implicit expectation about the outcome
and rely on a few ``directional signs'' to check when the outcome
deviate from those expectation. However, these expectations are rarely
made explicit within the analysis workflow. This makes it challenging
for consumers of the analysis to evaluate the results, since it becomes
difficult to disentangle whether discrepancies arise from differing
expectations or from the use of statistical technique, without running
the analysis themselves. Non-expert analysts, lacking prior knowledge or
instinct, may not have clear expectations of the results. This can lead
to reduced confidence of the analysis and makes it more difficult and
time-consuming to diagnose the cause of the deviation when the results
don't align with expectations.

By explicitly formulating these expectations, an analysis plan can guide
the analysis process, facilitate the communication and comparison with
other analyses and evaluate the validity of the results.

@@

  \begin{itemize}
  \tightlist
  \item
    and can be used to guide the analysis process
  \item
    with simulation, the unit tests can be used to construct the
    ``result universe''
  \item
    With formulated expectation, one can generate multiple version of
    the dataset for teaching statistics
  \end{itemize}
\end{itemize}

### 2024-09-13 index.qmd
more definition and explanation on expectations and the current gap

* describe/ define what analysis plan is, expectation (outcome and plan)
    * expectations can be used to formulate unit tests, they can be used to guide to diagnose in the analysis process: data issue, unreasonable expectation -- expectation is "wrong" 
    * some expectations can be about data: e.g. check temporal gaps in time series data. Some others are statistical checks: sample size, outliers, model specification, or checks based on domain knowledge: include temperature in the mortality ~ PM10 model.

@@

An analysis plan is a set of analysis steps plus their expected outcomes. An analysis plan consists of two parts: outcome expectation and plan expectation. Outcome expectation is the expected result of the analysis, for example, a significant p-value or a coefficient estimate within a certain range. Plan expectation is the expected results from the intermediate steps of the analysis, for example, the expected distribution of the data, or the expected correlation structure between variables.

While there is typically one major result expectation in an analysis, multiple plan expectations can be formulated as unit tests to guide the analysis process. These plan expectations serve as road marks to test, at various stages of the analysis, whether the results are within expectation. When the outcome is out of expectation, these plan expectations can serve as unit tests to diagnose the analysis process.

* why it is useful to be explicit about expectation? 
    * with simulation, the unit tests can be used to construct the "result universe" 
    * With formulated expectation, one can generate multiple version of the dataset for teaching statistics

>>>>

@@

TODO: more about analysis plan itself here

TODO: what if the expectation is "wrong"

An analysis plan is a set of analysis steps combined with expectations. Expectations represent our belief about certain aspects of the analysis, independent of the analysis itself. It can be divided into two types: *outcome expectation* and *plan expectation*. Outcome expectation refers to what we anticipate from the main result of the analysis based on prior knowledge. They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates [@grolemund_cognitive_2014]. For example, in public health, prior research shows the average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient]. This serves as an expectation for similar future studies. Plan expectations concern the intermediate steps within the analysis rather than the final outcome. They serve as checkpoints to detect deflection in the analysis process For example, we may expect temporal data to be ordered with no gaps and duplicates, or expect that temperature will be a significant covariate in the linear regression model of PM10 on mortality.

(might be useful) Analysis plans can be constructed at various granularities, at the highest level, one may only has a plan of the specific method used for analysing data and the expected outcome. This provides little guidance when a deviation from expectation occurs. At the lowest level, one may have a plan for each data entry and every data handling steps. This provides too much detail and may not be practical in practice.

Experienced analysts often have implicit expectation about the outcome and rely on a few "directional signs" to check when the outcome deviate from those expectation. However, these expectations are rarely made explicit within the analysis workflow. This makes it challenging for consumers of  the analysis to evaluate the results, since it becomes difficult to disentangle whether discrepancies arise from differing expectations or from the use of statistical technique, without running the analysis themselves. Non-expert analysts, lacking prior knowledge or instinct, may not have clear expectations of the results. This can lead to reduced confidence of the analysis and makes it more difficult and time-consuming to diagnose the cause of the deviation when the results don't align with expectations.

By explicitly formulating these expectations, an analysis plan can guide the analysis process, facilitate the communication and comparison with other analyses and evaluate the validity of the results.

* why it is useful to be explicit about expectation? 
    * and can be used to guide the analysis process
    * with simulation, the unit tests can be used to construct the "result universe" 
    * With formulated expectation, one can generate multiple version of the dataset for teaching statistics
